---
title: "Quantium DA Task1"
author: "Santosh Reddy Edulapalle"
date: "2023-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the datasets

```{r}
#install.packages('readxl')
library('readxl')
trans <- read_excel('QVI_transaction_data.xlsx')
```

Examine transaction data

```{r}
#showing head ( top 10 rows)
head(trans)
```
```{r}
#showing summary
summary(trans)
```
```{r}
#showing high level structure
str(trans)
```

We can see that the DATE type is DOUBLE We need to convert it to DATE type

```{r}
typeof(trans$DATE)
trans$DATE <- as.Date(trans$DATE,origin = '1899-12-30')
typeof(trans$DATE)
#examine structure
str(trans)
```

Checking PROD_NAME
Since PROD_NAME is a name given to individual object, we will factorise it and make them into groups.

```{r}
trans$PROD_NAME_FACTOR <- factor(trans$PROD_NAME)
summary(trans$PROD_NAME_FACTOR)

```

```{r}
summary(trans)
str(trans)
```


Examine product words in PROD_NAME

```{r}
library(data.table)
productWords <- data.table(unlist(strsplit(unique(trans$PROD_NAME), " ")))
setnames(productWords, 'words')
#productWords
```

Removing words that contain numericals

```{r}
numerical.validation <- grepl('[1-9]',productWords[,words])
productWords <- productWords[numerical.validation==FALSE]
#productWords
```

Removing words that contain special character '&'

```{r}
scAnd.validation <- grepl('&',productWords[,words])
productWords <- productWords[scAnd.validation==FALSE]
```

Removing words that contain special character '/'

```{r}
sc.validation <- grepl('/',productWords[,words])
productWords <- productWords[sc.validation==FALSE]
```

Counting frequencies

```{r}
#factorising words
productWords <- factor(productWords$words)
```

summary

```{r}
summary(productWords)
```
```{r}
trans <- data.table(trans)
```

```{r}
# Remove salsa products
trans[, SALSA := grepl("salsa", tolower(PROD_NAME))]
trans <- trans[SALSA == FALSE, ][, SALSA := NULL]
```

checking summary

```{r}
summary(trans)
```

checking for outliers : By seeing summary of data, we can see that the max value of PROD_QTY is more that 3rd quartile+ 1.5 IQR


Lets confirm this with a boxplot

```{r}
boxplot(trans$PROD_QTY)
```

Yes we can confirm existance of outliers.

```{r}
library(dplyr)
filter(trans,trans$PROD_QTY==200)
```

We have 2 records where the PROD_QTY is 200. Both are made by same customer 226000.
Let's see if he has any other transactions

```{r}
filter(trans,trans$LYLTY_CARD_NBR==226000)

```

Removing customer - 226000 from further analysis

```{r}
#trans[,trans$LYLTY_CARD_NBR != 226000]
trans <- trans[trans[,trans$LYLTY_CARD_NBR != 226000]]
```


Checking summary 

```{r}
summary(trans)
```

```{r}
boxplot(trans$PROD_QTY)
```

Let us factorise the dates

```{r}
trans$newDATE <- factor(trans$DATE)
```

Summary
```{r}
summary(trans)
str(trans)
```

There are 364 unique dates where transaction happend.
We will create a new column with dates from min to max i.e., 2018-07-01 to 2019-06-30
and then join this with trans to find that missing date.

```{r}
model.date <- seq(as.Date("2018-07-01"),as.Date("2019-06-30"),by = 'day')
model.date <- data.table(model.date)
setnames(model.date,'DATE')
#colnames(model.date) <- c('Date')
```

joining

```{r}

trans <- full_join(trans,model.date,by = c('DATE'))
```

summary

```{r}
summary(trans)
```

finding the date

```{r}

filter(trans,is.na(trans$STORE_NBR) == TRUE)
```

We can see that the date 2018-12-25 is missing.


Counting number of transactions per date.

```{r}
transactions_per_date <- trans[, as.Date(trans$DATE, format = "%Y-%m-%d")]
transactions_per_date <- table(transactions_per_date)
transactions_per_date <- data.table(transactions_per_date)
```


Plotting


```{r}
library(ggplot2)
# Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
# Plot transactions over time
ggplot(transactions_per_date, aes(x = as.Date(transactions_per_date), y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see some anomaly December.
Creating December chart to further investigate.

```{r}
x = subset(transactions_per_date, format.Date(transactions_per_date,"%m")=="12")
ggplot(x, aes(x = as.Date(transactions_per_date), y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 day") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that on Dec 25 we do not have any transaction. Because it was a missing value.
Sales got increased until Christmas day and on Christmas day shops were closed.

```{r}
# removing the Christmas day
trans <- subset(trans,trans$DATE != '2018-12-25')
```

Creating new features: Pack size

```{r Create pack size}
library(readr)
# creating new Pack size feature in trans by parasing our numbers from product names.
trans[, PACK_SIZE := parse_number(PROD_NAME)]
# Always check your output
# Let's check if the pack sizes look sensible
#.N is a spl variable in data.table used to represent # of observations in a group along with by = pack_size
trans[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

Histogram of Pack size.

```{r}
x1 <- trans$PACK_SIZE
x1 <- table(x1)
x1<- data.table(x1)
colnames(x1) <- c('Pack_size','Transactions')
```

```{r}
barplot(height = x1$Transactions,
        names.arg = x1$Pack_size,
        main="Histogram of Pack_size to Transactions",
        xlab = "Pack size", 
        ylab= "# Transactions")
```

Creating brand names

```{r}
#Here we are parsing the first word of the sentence using word() from stringr
trans$Brand_name <- word(trans$PROD_NAME, 1)
```

```{r}
#checking brands results
trans[, .N, Brand_name][order(Brand_name)]
```

Some of the brand names look like they are of the same brands - such as RED and
RRD, NCC and Natural Chip Co, Smith and Smiths, infuzions and infzns, Snbts and Sunbites, WW and Woolworths, Dorito and Doritos, Grain and GrnWves
Let's combine these together.

```{r}
#clean brand names
trans[Brand_name == "Red", Brand_name := "RRD"]
trans[Brand_name == "Dorito", Brand_name := "Doritos"]
trans[Brand_name == "GrnWves", Brand_name := "Grain Waves"]
trans[Brand_name == "Grain", Brand_name := "Grain Waves"]
trans[Brand_name == "Natural", Brand_name := "NCC"]
trans[Brand_name == "Smith", Brand_name := "Smiths"]
trans[Brand_name == "Infzns", Brand_name := "Infuzions"]
trans[Brand_name == "Snbts", Brand_name := "Sunbites"]
trans[Brand_name == "Woolworths", Brand_name := "WW"]

```

```{r}
#checking brands results
trans[, .N, Brand_name][order(Brand_name)]
```



