---
title: "Quantium DA Task1"
author: "Santosh Reddy Edulapalle"
date: "2023-02-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load required libraries
```{r}
library(readxl)
library(data.table)
library(ggplot2)
library(ggmosaic)
library(readr)
library(stringr)
library(dplyr)
library(arules)
library(arulesViz)
```

#### Loading the dataset

```{r}

trans <- read_excel('QVI_transaction_data.xlsx')
```

## Exploratory data analysis on Transaction data

The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.

#### Examine transaction data

```{r}
#showing head ( top 10 rows)
head(trans)
```
```{r}
#showing summary
summary(trans)
```
```{r}
#showing high level structure
str(trans)
```
#### Convert DATE column to a date format
We can see that the DATE type is DOUBLE We need to convert it to DATE type
CSV and Excel integer dates begin on 30 Dec 1899
```{r}
typeof(trans$DATE)
trans$DATE <- as.Date(trans$DATE,origin = '1899-12-30')
typeof(trans$DATE)
#examine structure
str(trans)
```

#### Examine PROD_NAME
Since PROD_NAME is a name given to individual object, we will factorise it and make them into groups.

```{r}
trans$PROD_NAME_FACTOR <- factor(trans$PROD_NAME)
summary(trans$PROD_NAME_FACTOR)

```

```{r}
summary(trans)
str(trans)
```

#### Text analysis

Looks like we are definitely looking at potato chips but how can we check that these are all chips? We can
do some basic text analysis by summarising the individual words in the product name.

#### Examine product words in PROD_NAME

```{r}

productWords <- data.table(unlist(strsplit(unique(trans$PROD_NAME), " ")))
setnames(productWords, 'words')

```

As we are only interested in words that will tell us if the product is chips or not, let’s remove all words with
digits and special characters such as ‘&’ from our set of product words. We can do this using grepl().

#### Removing words that contain numerical

```{r}
numerical.validation <- grepl('[1-9]',productWords[,words])
productWords <- productWords[numerical.validation==FALSE]
#productWords
```

#### Removing words that contain special character '&'

```{r}
scAnd.validation <- grepl('&',productWords[,words])
productWords <- productWords[scAnd.validation==FALSE]
```

#### Removing words that contain special character '/'

```{r}
sc.validation <- grepl('/',productWords[,words])
productWords <- productWords[sc.validation==FALSE]
```

#### Counting frequencies

```{r}
#factorising words
productWords <- factor(productWords$words)
```

summary

```{r}
summary(productWords)
```

```{r}
trans <- data.table(trans)
```
There are salsa products in the dataset but we are only interested in the chips category, so let’s remove
these.
#### Remove salsa products
```{r}
# Remove salsa products
trans[, SALSA := grepl("salsa", tolower(PROD_NAME))]
trans <- trans[SALSA == FALSE, ][, SALSA := NULL]
```

#### Summarise the data to check for nulls and possible outliers

```{r}
summary(trans)
```

#### Checking for outliers
By seeing summary of data, we can see that the maximum value of PROD_QTY is more that (3rd quartile + 1.5*IQR)


Lets confirm this with a boxplot

```{r fig.width = 10, fig.align = "center"}
boxplot(trans$PROD_QTY)
```

Yes we can confirm existence of outliers.

Let’s investigate further the case where 200 packets of chips are bought in one transaction.
```{r}

filter(trans,trans$PROD_QTY==200)
```

We have 2 records where the PROD_QTY is 200. Both are made by same customer 226000.

Let's see if he has any other transactions

```{r}
filter(trans,trans$LYLTY_CARD_NBR==226000)

```

It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.

#### Filter out the customer based on the loyalty card number
Removing customer - 226000 from further analysis

```{r}
#trans[,trans$LYLTY_CARD_NBR != 226000]
trans <- trans[trans[,trans$LYLTY_CARD_NBR != 226000]]
```

#### Re-examine transaction data


```{r}
summary(trans)
```

```{r fig.width = 10, fig.align = "center"}
boxplot(trans$PROD_QTY)
```

#### Count the number of transactions by date
Let us factorise the dates

```{r}
trans$newDATE <- factor(trans$DATE)
```

Summary
```{r}
summary(trans)
str(trans)
```

There are 364 unique dates where transaction happened.
We will create a new column with dates from min to max i.e., 2018-07-01 to 2019-06-30
and then join this with trans to find that missing date.

```{r}
model.date <- seq(as.Date("2018-07-01"),as.Date("2019-06-30"),by = 'day')
model.date <- data.table(model.date)
setnames(model.date,'DATE')
#colnames(model.date) <- c('Date')
```

Full-join Date

```{r}

trans <- full_join(trans,model.date,by = c('DATE'))
```

summary

```{r}
summary(trans)
```

finding the date

```{r}

filter(trans,is.na(trans$STORE_NBR) == TRUE)
```

We can see that the date 2018-12-25 is missing.


#### Count the number of transactions by date

```{r}
transactions_per_date <- trans[, as.Date(trans$DATE, format = "%Y-%m-%d")]
transactions_per_date <- table(transactions_per_date)
transactions_per_date <- data.table(transactions_per_date)
```


#### Plot transactions over time


```{r fig.width = 10, fig.align = "center"}

#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(transactions_per_date, aes(x = as.Date(transactions_per_date), y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

#### Filter to December and look at individual days
We can see some anomaly in December.
Creating December chart to further investigate.

```{r fig.width = 10, fig.align = "center"}
x = subset(transactions_per_date, format.Date(transactions_per_date,"%m")=="12")
ggplot(x, aes(x = as.Date(transactions_per_date), y = N)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 day") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We can see that on Dec 25 we do not have any transaction. Because it was a missing value.
Sales got increased until Christmas day and on Christmas day shops were closed.

```{r}
# removing the Christmas day
trans <- subset(trans,trans$DATE != '2018-12-25')
```

## Feature Engineering

Now that we are satisfied that the data no longer has outliers, we can move on to creating other features
such as brand of chips or pack size from PROD_NAME. We will start with pack size.

#### Creating new features: Pack size

```{r Create pack size}

# creating new Pack size feature in trans by parasing our numbers from product names.
trans[, PACK_SIZE := parse_number(PROD_NAME)]
# Always check your output
# Let's check if the pack sizes look sensible
#.N is a spl variable in data.table used to represent # of observations in a group along with by = pack_size
trans[, .N, PACK_SIZE][order(PACK_SIZE)]
```

The largest size is 380g and the smallest size is 70g - seems sensible!

#### Histogram of Pack size.

```{r fig.width = 10, fig.align = "center"}
x1 <- trans$PACK_SIZE
x1 <- table(x1)
x1<- data.table(x1)
colnames(x1) <- c('Pack_size','Transactions')

barplot(height = x1$Transactions,
        names.arg = x1$Pack_size,
        main="Histogram of Pack_size to Transactions",
        xlab = "Pack size", 
        ylab= "# Transactions")
```

#### Creating new features: Brand_name
Pack sizes created look reasonable and now to create brands, we can use the first word in PROD_NAME to
work out the brand name

```{r}
#Here we are parsing the first word of the sentence using word() from stringr
trans$Brand_name <- word(trans$PROD_NAME, 1)

#checking brands results
trans[, .N, Brand_name][order(Brand_name)]
```

#### Clean brand names
Some of the brand names look like they are of the same brands - such as RED and
RRD, NCC and Natural Chip Co, Smith and Smiths, infuzions and infzns, Snbts and Sunbites, WW and Woolworths, Dorito and Doritos, Grain and GrnWves
Let's combine these together.

```{r}
#clean brand names
trans[Brand_name == "Red", Brand_name := "RRD"]
trans[Brand_name == "Dorito", Brand_name := "Doritos"]
trans[Brand_name == "Grain", Brand_name := "GrnWves"]
trans[Brand_name == "NCC", Brand_name := "Natural"]
trans[Brand_name == "Smith", Brand_name := "Smiths"]
trans[Brand_name == "Infzns", Brand_name := "Infuzions"]
trans[Brand_name == "Snbts", Brand_name := "Sunbites"]
trans[Brand_name == "WW", Brand_name := "Woolworths"]

```

```{r}
#checking brands results
trans[, .N, Brand_name][order(Brand_name)]
```



#### Loading customer dataset

```{r load customer data}
cust = read.csv('QVI_purchase_behaviour.csv')
```

### Examining customer data
```{r}
summary(cust)
str(cust)
```

Let’s have a closer look at the LIFESTAGE and PREMIUM_CUSTOMER columns.
```{r fig.width = 10, fig.align = "center"}
#distribution of lifestage and premium_customer
ggplot(data = cust, aes(x = LIFESTAGE)) +
  geom_histogram(stat = "count", fill = "#56B4E9") +
  geom_text(stat = "count", aes(label = ..count..), fontface = "bold", vjust = 5)

ggplot(data = cust,aes(x = PREMIUM_CUSTOMER)) + 
  geom_histogram(stat = "count", fill = "#E69F00") + 
  geom_text(stat = "count",aes(label = ..count..), fontface = "bold",vjust = 5)
```

```{r}

#### Examining the values of lifestage and premium_customer(same as above but just numerical values)
cust %>% 
  group_by(LIFESTAGE) %>%
  summarise(cust_count = n()) %>%
  arrange(desc(cust_count))

cust %>%
  group_by(PREMIUM_CUSTOMER) %>%
  summarise(prem_count = n()) %>%
  arrange(desc(prem_count))
```

As there do not seem to be any issues with the customer data, we can now go ahead and join the transaction
and customer data sets together

#### Merge transaction data to customer data

```{r }
#### Merge transaction data to customer data
# all.x = T implies full left join
df <- merge(trans, cust, all.x = TRUE)

#df$LYLTY_CARD_NBR <- as.factor(df$LYLTY_CARD_NBR)
```

As the number of rows in `data` is the same as that of `transactionData`, we can be
sure that no duplicates were created. This is because we created `data` by setting
`all.x = TRUE` (in other words, a left join) which means take all the rows in
`transactionData` and find rows with matching values in shared columns and then
joining the details in these rows to the `x` or the first mentioned table.

Checking if any transactions did not have a matched customer.
```{r Check for missing customer details}
df[is.null(LIFESTAGE)]
df[is.null(PREMIUM_CUSTOMER)]
```

Great, there are no nulls! So all our customers in the transaction data has been accounted for in the customer
dataset.

```{r}
#Not adding DEC 25
df <- df[!DATE == '2018-12-25']
```

#### Saving dataset
```{r Code to save dataset as a csv}
filePath <- "/Users/santosh/Documents/QuantiumDA/quantiumDA/"
fwrite(df, paste0(filePath,"QVI_data.csv"))
```

Data exploration is now complete!

## Exploratory data analysis on customer segments

Now that the data is ready for analysis, we can define some metrics of interest to the client:
• Who spends the most on chips (total sales), describing customers by lifestage and how premium their
general purchasing behavior is
• How many customers are in each segment
• How many chips are bought per customer by segment
• What’s the average chip price by customer segment
We could also ask our data team for more information. Examples are:
• The customer’s total spend over the period and total spend for each transaction to understand what
proportion of their grocery spend is on chips
• Proportion of customers in each customer segment overall to compare against the mix of customers
who purchase chips.

Let’s start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and plotting the split by
these segments to describe which customer segment contribute most to chip sales.

#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER

```{r}
#salesxxx <- df[, .(SALES = sum(TOT_SALES)), .(LIFESTAGE, PREMIUM_CUSTOMER)]
#same method different approach, upper uses data.table functions, below uses dplyr package

sales_summary <- df %>%
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>%
  summarise(total_sales = sum(TOT_SALES))

```


```{r fig.width = 10, fig.align = "center"}

ggplot(sales_summary, aes(x = LIFESTAGE, y = total_sales, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = round(total_sales)), position = position_dodge(width = 0.9),fontface = "bold", vjust = -0.1, size = 3) +
  labs(title = "Total Sales by LIFESTAGE and PREMIUM_CUSTOMER") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```

Sales are coming mainly from Budget - older families, Mainstream - young
singles/couples, and Mainstream - retirees

Let’s see if the higher sales are due to there being more customers who buy chips.

#### Number of customers by LIFESTAGE and PREMIUM_CUSTOMER

```{r}
#count distinct values of customers
cust_summary <- df %>%
  distinct(LYLTY_CARD_NBR, .keep_all = TRUE) %>%
  count(LIFESTAGE, PREMIUM_CUSTOMER, name = "cust_count")

```

```{r fig.width = 10, fig.align = "center"}

ggplot(cust_summary, aes(x = LIFESTAGE, y = cust_count, fill = PREMIUM_CUSTOMER)) +
  geom_bar(stat = "identity", position = "dodge")  + 
  geom_text(aes(label = round(cust_count)), position = position_dodge(width = 0.9),fontface = "bold", vjust = -0.1, size = 3) +
  labs(title = "Number of Unique Customers by LIFESTAGE and PREMIUM_CUSTOMER") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.
Let's have a look at this next.

#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER

```{r}
avg_cust <- df %>% group_by(LIFESTAGE , PREMIUM_CUSTOMER ) %>% 
  summarize(avg_cust_count = sum(PROD_QTY)/n_distinct(LYLTY_CARD_NBR))
```

```{r fig.width = 10, fig.align = "center"}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(data = avg_cust,aes(x = LIFESTAGE,y = avg_cust_count,fill = PREMIUM_CUSTOMER)) +   geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label = round(avg_cust_count)),fontface = "bold",vjust = -0.1, size = 3, position = position_dodge(width = 0.9))+
  labs(title = "Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Older families and young families in general buy more chips per customer.

Let's also investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.

#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
```{r}
avg_ppu <- df %>% group_by(LIFESTAGE , PREMIUM_CUSTOMER) %>%
  summarise(avg_price = (sum(TOT_SALES)/sum(PROD_QTY)))
```



```{r fig.width = 10, fig.align = "center"}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
ggplot(data = avg_ppu,aes(x = LIFESTAGE,y = avg_price,fill = PREMIUM_CUSTOMER)) +   geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(avg_price,2)), position = position_dodge(width = 0.9),fontface = "bold",vjust = -0.1,size = 3) +
  labs(title = "Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Mainstream midage and young singles and couples are more willing to pay more per
packet of chips compared to their budget and premium counterparts. This may be due
to premium shoppers being more likely to buy healthy snacks and when they buy
chips, this is mainly for entertainment purposes rather than their own consumption.
This is also supported by there being fewer premium midage and young singles and
couples buying chips compared to their mainstream counterparts.

#### Perform an independent t-test

Here we are performing t-test between Mainstream vs (Premium or Budget) for Midage and young - singles and couples.

```{r t-test: mainstream vs premium}
#### Perform an independent t-test between mainstream vs premium and budget midage and young singles and couples
# Over to you! Perform a t-test to see if the difference is significant.
main_premium <- subset(avg_ppu, PREMIUM_CUSTOMER %in% c("Mainstream", "Premium")&LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES"))

t.test(avg_price~PREMIUM_CUSTOMER, data = main_premium)
```

```{r t-test: mainstream vs budget}
#### Perform an independent t-test between mainstream vs premium and budget midage and young singles and couples
# Over to you! Perform a t-test to see if the difference is significant.
main_budget <- subset(avg_ppu, PREMIUM_CUSTOMER %in% c("Mainstream","Budget")&LIFESTAGE %in% c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES"))

t.test(avg_price~PREMIUM_CUSTOMER, data = main_budget)
```

```{r t-test: mainstream vs others}

t.test(avg_ppu[avg_ppu$LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") & avg_ppu$PREMIUM_CUSTOMER == "Mainstream", "avg_price"],
       avg_ppu[avg_ppu$LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") & avg_ppu$PREMIUM_CUSTOMER != "Mainstream", "avg_price"],
       alternative = "greater")

```


All the t-test results suggest that p-value is less than $\alpha = 0.05$ and we accept alternate hypothesis that there is some difference between mean between the groups mainstream vs premium or budget.

the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples.


#### Deep dive into Mainstream, young singles/couples 

We might want to target customer segments that contribute the most to sales to
retain them or further increase sales. Let's look at Mainstream - young
singles/couples. For instance, let's find out if they tend to buy a particular
brand of chips.

```{r}
#### Deep dive into Mainstream, young singles/couples
segment1 <- df[LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream",]
other <- df[!(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream"),]
#### Brand affinity compared to the rest of the population
quantity_segment1 <- segment1[, sum(PROD_QTY)]
quantity_other <- other[, sum(PROD_QTY)]
quantity_segment1_by_brand <- segment1[, .(targetSegment = sum(PROD_QTY)/quantity_segment1), by = Brand_name]
quantity_other_by_brand <- other[, .(other = sum(PROD_QTY)/quantity_other), by = Brand_name]
brand_proportions <- merge(quantity_segment1_by_brand, quantity_other_by_brand)[, affinityToBrand := targetSegment/other]
brand_proportions[order(-affinityToBrand)]
```

We can see that :
• Mainstream young singles/couples are 22% more likely to purchase Tyrrells chips compared to the
rest of the population
• Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest
of the population

Let’s also find out if our target segment tends to buy larger packs of chips.

#### Preferred pack size compared to the rest of the population

```{r}
#### Preferred pack size compared to the rest of the population
quantity_segment1_by_pack <- segment1[, .(targetSegment = sum(PROD_QTY)/quantity_segment1), by = PACK_SIZE]
quantity_other_by_pack <- other[, .(other = sum(PROD_QTY)/quantity_other), by = PACK_SIZE]
pack_proportions <- merge(quantity_segment1_by_pack, quantity_other_by_pack)[, affinityToPack := targetSegment/other]
pack_proportions[order(-affinityToPack)]
```

It looks like Mainstream young singles/couples are 27% more likely to purchase a 270g pack of chips compared
to the rest of the population but let’s dive into what brands sell this pack size.

```{r}
df[PACK_SIZE == 270, unique(PROD_NAME)]

df[Brand_name == "Tyrrells", unique(PROD_NAME)]
```

Twisties are the only brand offering 270g packs and so this may instead be reflecting a higher likelihood of
purchasing Twisties.

Even though Tyrrells is top likely brand for Mainstream - young singles/couples, they are 11% less likely to purchase 165g packets.

#### Conclusion

Sales have mainly been due to Budget - older families, Mainstream - young singles/couples, and Mainstream
- retirees shoppers. We found that the high spend in chips for mainstream young singles/couples and retirees
is due to there being more of them than other buyers. Mainstream, midage and young singles and
couples are also more likely to pay more per packet of chips. This is indicative of impulse buying behaviour.
We’ve also found that Mainstream young singles and couples are 23% more likely to purchase Tyrrells chips
compared to the rest of the population. The Category Manager may want to increase the category’s performance
by off-locating some Tyrrells and smaller packs of chips in discretionary space near segments
where young singles and couples frequent more often to increase visibilty and impulse behaviour.

With these Insights, my task-1 is completed.

#### End of Task-1